Extracting OSM data

### 2023-02-28 data-analytics.qmd
data prep step: extract runway information for study from OSM

```{r}
aip_apts <- tribble(
  ~ICAO,   ~REF,         ~LAT,    ~LON,     ~ELEV, ~SRC 
  ,"EGLL", "OSN-center", 51.4706, -0.461941, 83, "https://opensky-network.org/airport-profile?icao=EGLL"
  ,"EGLL", "ARP",        51.4775, -0.461389, NA, "AIP"
  ,"EGLL", "THR-09L",    51.4775, -0.484992, 79 ,"AIP - 512839.00N 0002905.97W THR 78.6 FT"
  ,"EGKK", "OSN-center", 51.148102,-0.190277, 202,"https://opensky-network.org/airport-profile?icao=EGKK"
  ,"EDDF", "OSN-center", 50.033333,8.570556, 364,"https://opensky-network.org/airport-profile?icao=EDDF"
  ,"EHAM", "OSN-center", 52.308601,4.763889, -11, "https://opensky-network.org/airport-profile?icao=EHAM"
  ,"KATL", "OSN-center", 33.6366996,-84.4281005,1026,"https://opensky-network.org/airport-profile?icao=KATL"
  ,"KORD", "OSN-center", 41.9785996,-87.9048004,672, "https://opensky-network.org/airport-profile?icao=KORD"
  ,"KJFK", "OSN-center", 40.6398010,-73.7789002, 13, "https://opensky-network.org/airport-profile?icao=KJFK"
)
write_csv(aip_apts, file = "./data/aip.csv")
aip_apts
```

@@

* trajectory of LAT/LON within a certain distance around an airport and below a certain altitude

>>>>

```{r}
aip_apts <- tribble(
  ~ICAO,   ~REF,         ~LAT,    ~LON,     ~ELEV, ~SRC 
  ,"EGLL", "OSN-center", 51.4706, -0.461941, 83, "https://opensky-network.org/airport-profile?icao=EGLL"
  ,"EGLL", "ARP",        51.4775, -0.461389, NA, "AIP"
  ,"EGLL", "THR-09L",    51.4775, -0.484992, 79 ,"AIP - 512839.00N 0002905.97W THR 78.6 FT"
  ,"EGKK", "OSN-center", 51.148102,-0.190277, 202,"https://opensky-network.org/airport-profile?icao=EGKK"
  ,"EDDF", "OSN-center", 50.033333,8.570556, 364,"https://opensky-network.org/airport-profile?icao=EDDF"
  ,"EHAM", "OSN-center", 52.308601,4.763889, -11, "https://opensky-network.org/airport-profile?icao=EHAM"
  ,"KATL", "OSN-center", 33.6366996,-84.4281005,1026,"https://opensky-network.org/airport-profile?icao=KATL"
  ,"KORD", "OSN-center", 41.9785996,-87.9048004,672, "https://opensky-network.org/airport-profile?icao=KORD"
  ,"KJFK", "OSN-center", 40.6398010,-73.7789002, 13, "https://opensky-network.org/airport-profile?icao=KJFK"
)
#write_csv(aip_apts, file = "./data/aip.csv")
aip_apts
```

@@

* trajectory of LAT/LON within a certain distance around an airport and below a certain altitude
* we start with Heathrow, filtering for some arrivals

> ds |> filter(grepl(pattern = "^BAW", x = FLTID ), ALT_B < 1000) |> distinct(FLTID)
# A tibble: 761 × 1
   FLTID  
   <chr>  
 1 BAW112B
 2 BAW32  
 3 BAW237 
 4 BAW236 
 5 BAW16  
 6 BAW28  
 7 BAW34  
 8 BAW56  
 9 BAW262 
10 BAW74

```{r}
baws <- ds |> filter(grepl(pattern = "^BAW", x = FLTID )) |> 
  group_by(FLTID) |> 
  mutate(
    MST = case_when(
          TIME == min(TIME) ~ "START"
          ,TIME == max(TIME) ~ "END")
    )
```
```{r}
picks <- c("BAW112B", "BAW236","BAW56","BAW262","BAW74")

ggplot() +
  geom_point(data = aip_apts |> filter(ICAO == "EGLL")
             , aes(x = LON, y = LAT), size = 3, color = "blue") +
  geom_point(data = baws |> filter(FLTID %in% picks)
             , aes(x = LON, y = LAT, color = MST))
```

### 2023-02-28 x_osm-extract-rwys.qmd
data prep step: extract runway information for study from OSM

@@

```{r}
extract_rwy_linestrings <- function(.osm_sf){
  rwys_ls <- .osm_sf$osm_lines |> filter(aeroway == "runway", is.na(runway))
  message("number of runways: ", nrow(rwys_ls))
  rwys_ls
}

@@

extract_thresholds <- function(.rwy_ls){
  thr1 <- lwgeom::st_startpoint(.rwy_ls) |> pt_to_latlon()
  thr2 <- lwgeom::st_endpoint(  .rwy_ls) |> pt_to_latlon()

  bind_rows(thr1, thr2)
}
```

@@

>>>>

Development version

Data preparatory step: extract runway information from Openstreet Map (OSM).

@@

```{r}
extract_rwy_linestrings <- function(.osm_sf){

 osm_values <- c("osm_id", "aeroway", "ref", "runway", "geometry")

  this_rwys_ls <- .osm_sf$osm_lines |>
    dplyr::select(any_of(osm_values)) 
  message("missing values: ", setdiff(osm_values, names(this_rwys_ls)))

  # fix EDDF no runway-tag!
  if("runway" %in% setdiff(osm_values, names(this_rwys_ls))){
    this_rwys_ls <- this_rwys_ls |> mutate(runway = NA_character_)
  }

  this_rwys_ls <- this_rwys_ls |> 
    dplyr::filter(aeroway == "runway", is.na(runway))
  message("number of runways: ", nrow(this_rwys_ls))
  this_rwys_ls
}

@@

extract_thresholds <- function(.rwys_ls, .drop_geometry = FALSE){
  rwys <- .rwys_ls |> select(aeroway, ref, geometry)
  thr1 <- lwgeom::st_startpoint(.rwys_ls) |> pt_to_latlon() |> rename(THR1_LAT = LAT, THR1_LON = LON)
  thr2 <- lwgeom::st_endpoint(  .rwys_ls) |> pt_to_latlon() |> rename(THR2_LAT = LAT, THR2_LON = LON)

  bind_cols(rwys, thr1, thr2)  # todo: add drop_geometry
}
```

@@

get thresholds

* sfnetworks has edge_azimuth() function, coerce bearing into degrees
* calculate (rough) offset between bearing and runway identifier & select

```{r}
rq <-  extract_thresholds(rwys_ls)

rq2 <- rq |> 
  sfnetworks::as_sfnetwork() |> 
  sfnetworks::activate("edges") |> 
  dplyr::mutate(  
      BRG = sfnetworks::edge_azimuth()
    , BRG = units::set_units(BRG, "degrees")
    , BRG = (as.numeric(BRG) + 360) %% 360

    , REF1 = stringr::str_split_i(ref, pattern = "/",1)
    , REF2 = stringr::str_split_i(ref, pattern = "/",2)

    , RWY_1 = if_else( abs( (parse_number(REF1) * 10) - BRG ) < 30, REF1, REF2)
    , RWY_2 = if_else( abs( (parse_number(REF2) * 10) - BRG ) > 30, REF2, REF1)
         )
rq2
```

Package output

```{r}
apt <- "EGKK"

# pull out edges/runways 
rq3 <- rq2 |> sfnetworks::activate("edges") |> st_as_sf() |> sf::st_drop_geometry()

rwy_ends <-
  bind_cols(rq3 |> select(REF = RWY_1, LAT = THR1_LAT, LON = THR1_LON, NAME = ref)) |> 
  bind_rows(rq3 |> select(REF = RWY_2, LAT = THR2_LAT, LON = THR2_LON, NAME = ref)) |> 
  mutate(ICAO = apt) |> select(ICAO, everything())
rwy_ends
```

```{r}
ggplot() + 
  geom_sf(data = rwys_ls) +
  geom_point(data = rwy_ends, aes(x = LON, y = LAT)) +
  geom_text(data = rwy_ends, aes(x = LON, y = LAT, label = REF), vjust = -0.5)
```

Let's go for Heathrow

```{}
# fails from github! ### source("./R/xhelper_read_excel_from_url.R")
url   <- "https://github.com/euctrl-pru/airport-dashboard/blob/master/data/APT_BBOX.xlsx"
# github urls follow the pattern: https://raw.github.com/user/repository/branch/file.name
github_download_url <- "https://raw.github.com/euctrl-pru/airport-dashboard/master/data/APT_BBOX.xlsx"

## bboxs <- read_excel_from_url(url)
download.file(github_download_url, destfile = "./data/airport-bbox.xlsx", mode = "wb")
bboxs <- readxl::read_excel("./data/airport-bbox.xlsx")
```

```{r}
query_osm_bbox <- function(.bb_lonlat){
  q <- opq(bbox = .bb_lonlat) %>% 
    add_osm_feature(
      key = "aeroway"
      ,value =c("aerodrome", "apron", "control_tower", "gate", "hangar"
              ,"helipad", "runway", "taxiway", "terminal") ) %>% 
    osmdata_sf() 
}

plot_airport <- function(.osm_apt, .apt=NULL){
  g <- ggplot() +
  geom_sf(data = .osm_apt$osm_polygons
          ,inherit.aes = FALSE
          ,color = "lightblue"
          #,fill  = "lightblue"
          ) +
  geom_sf(data = .osm_apt$osm_lines %>% filter(aeroway != "runway")
          , color = "grey"
          ) +
  geom_sf(data = .osm_apt$osm_lines %>% filter(aeroway == "runway"),
          inherit.aes = FALSE,
          color = "black",
          size = 2 #.4
          ,alpha = .8) +
  theme_void() +
  labs(title = .apt)
 return(g)
}

coerce_bbox_string_to_numeric_vector <- function(.bbox_string){
  .bbox_string |> 
    str_split_1(pattern = ",") |> 
    as.numeric()
}
```

```{r}
egll_bb_lonlat <- bboxs |> 
  filter(AIRPORT == "EGLL") |> pull(BBOX) |> 
  coerce_bbox_string_to_numeric_vector()

osm_apt <- query_osm_bbox(egll_bb_lonlat)

plot_airport(osm_apt)
```

Check for runway ends

```{r}
rwys_ls <- extract_rwy_linestrings(osm_apt)
extract_thresholds(rwys_ls)
```

```{r}
associate_rwy_brg <- function(.thrs_ls){
  .thrs_ls |> 
  sfnetworks::as_sfnetwork() |> 
  sfnetworks::activate("edges") |> 
  dplyr::mutate(  
      BRG = sfnetworks::edge_azimuth()
    , BRG = units::set_units(BRG, "degrees")
    , BRG = (as.numeric(BRG) + 360) %% 360

    , REF1 = stringr::str_split_i(ref, pattern = "/",1)
    , REF2 = stringr::str_split_i(ref, pattern = "/",2)

    , RWY_1 = if_else( abs( (parse_number(REF1) * 10) - BRG ) < 30, REF1, REF2)
    , RWY_2 = if_else( RWY_1 == REF1, REF2, REF1)
         )
}
# rq2 <- rq |> associate_rwy_brg()
# rq2

pull_out_runway_info <- function(.thrs_brgs, .apt = apt){
# pull out edges/runways 
 tmp <- .thrs_brgs |> 
   sfnetworks::activate("edges") |> 
   sf::st_as_sf() |> 
   sf::st_drop_geometry()

 rwy_ends <-
  bind_cols(tmp |> select(REF = RWY_1, LAT = THR1_LAT, LON = THR1_LON, NAME = ref)) |> 
  bind_rows(tmp |> select(REF = RWY_2, LAT = THR2_LAT, LON = THR2_LON, NAME = ref)) |> 
  mutate(ICAO = .apt) |> select(ICAO, everything())
 rwy_ends
}
```

Executed for

* EGLL
* EHAM
* EGKK
* EDDF
* WSSS
* KATL, KORD, KJFK
* SBGR

```{r}
apt <- "SBGR"

# European airports
bboxs <- readxl::read_excel("./data/airport-bbox.xlsx")
# international airports
bboxs_int <- readxl::read_excel("./data/airport-intnl-bbox.xlsx")
bboxs <- bind_rows(bboxs, bboxs_int)

bb_lonlat <- bboxs |> 
  filter(AIRPORT == apt) |> pull(BBOX) |> 
  coerce_bbox_string_to_numeric_vector()

osm_apt <- query_osm_bbox(bb_lonlat)

rwys_ls <- extract_rwy_linestrings(osm_apt)

if(apt == "EDDF"){
  rwys_ls$ref <- gsub(pattern = "^18$", replacement = "18/36", x = rwys_ls$ref)
}

rwy_ends <-  rwys_ls |> 
  extract_thresholds() |> 
  associate_rwy_brg() |> 
  pull_out_runway_info()

write_csv(rwy_ends, file = paste0("./data/rwys-",apt,".csv"))
rwy_ends
```

Visual check

```{r}
plot_airport(osm_apt) +
  geom_point(data = rwy_ends, aes(x = LON, y = LAT)) +
  geom_text( data = rwy_ends, aes(x = LON, y = LAT, label = REF), vjust = -0.3)
```
